---
title: "Introduction to rslab"
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig.align: 'center'
    fig.asp: 0.618
    dpi: 320
    toc: true
    warning: FALSE
    message: FALSE
vignette: >
  %\VignetteIndexEntry{Introduction to rslab}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  dpi = 300,
  warning = FALSE,
  message = FALSE,
  out.width = "100%",
  comment = "#>"
)
```







## BTPS Calculator

[Spirometry test](https://en.wikipedia.org/wiki/Spirometry) is a procedure that measures lung volumes and air flow parameters which calculate from inspiratory and expiratory gas volumes of a human subject.

The gas volumes/flows obtained by the spirometer are in **ATPS** (Ambient Temperature and Pressure Saturated) conditions which its volumes also depends on environmental conditions including room temperature.

The obtained volumes/flows must be converted to gas condition in **BTPS** (Body Temperature, Pressure, water vapor Saturated) which gives more accurate representation of actual volumes/flows within the lungs.

\

[BTPS correction factor](https://nddmed.com/_Resources/Persistent/8a716ac3fd123ce0becd7b56596582d4fc4c0c47/appnote-btps-correction-v01r.pdf) is a coefficient used to convert flow and volume measured at ambient conditions (ATPS) to the conditions within the lungs (BTPS).

`get_btps_factor()` is a function to obtain the BTPS correction factor.

```{r setup}
library(rslab)
```

```{r btps}
# BTPS correction factor at room temperature = 27 degree celsius
btps_factor27 <- get_btps_factor(temp = 27)
btps_factor27
```

If Tidal Volume (TV) was measured as 500 ml at room temperature of 27 degree celsius, the 
TV in the lung at BTPS (body temperature, pressure, water vapor saturated) would be:

```{r}
# TV (ml) at BTPS
500 * btps_factor27 
```

If you want to convert several lung volumes or flow (from ATPS to BTPS), use `lung_vol_atps_btps()` and the result will be shown as a tibble.

```{r lung_vol_conv}
lung_vol_atps_btps(
  temp = 27,
  FEV1 = 3.5, 
  FVC = 4.5,
  PEF = 450,
  TV = 0.5,
  IC = 2.5,
  EC = 2.5,
  VC = 4.5
)
```

## Metabolic Rate & Oxygen Consumption Calculator

In a laboratory experiment using [Harvard Spirometer](https://www.somatco.com/Recording-Spirometer-50-1833-50-1817.pdf), this package provides functions to calculate metabolic rate and oxygen consumption by input a displacement in x-and y-direction of Harvard spirometer tracing, height and weight of the subject, and environmental conditions which are temperature and barometric pressure.

### Oxygen Consumption

In Harvard spirometer tracing, horizontal direction (x-direction) represents time (default paper speed = 25 mm/min) whereas vertical direction (y-direction) represents usage of oxygen (1 mm = 30 ml of oxygen). With a little bit of calculation, we can derived an oxygen consumption in unit of L/hr in the **ATPS** condition (**A**mbient **T**emperature and barometric **P**ressure **S**aturated with water vapor condition).

You can use `get_oxycons()` to calculate oxygen consumption and it will also provide other info printed to R console as well. 

```{r oxycons}
oxycons <- get_oxycons(x = 15, # displacement in x-direction = 15 mm
            y = 80, # displacement in y-direction = 80 mm
            paper_speed = 25 # Paper speed of the kymograph = 25 mm/min
            )
```


### Metabolic Rate

To calculate metabolic rate, we must first convert oxygen consumption $\dot{V}_{o_2}$ at **ATPS** to **STPD** (**S**tandard **T**emperature of 0°C and a barometric **P**ressure of 760 mmHg, and in a **D**ry state).

STPD correction factor can be used to convert gas in ATPS to BTPS condition.
It has a linear relationship with barometric pressure and temperature. 

Therefore, `get_STPD_factor()` can be used to predict STPD factor using `baro` and `temp_c` as predictors. (It use multiple linear regression model behind the scene.)


```{r stpd_760_25}
stpd_760_25 <- get_STPD_factor(
  baro = 760, # Barometric pressure at the recording site.
  temp_c = 25 # Temperature in celsius at the recording site.
  ) 
```

Correction can be made by multiply oxygen consumption $\dot{V}_{o_2}$ at ATPS to STPD correction factor.

```{r}
oxycons * stpd_760_25 # Unit in L/hr
```

Finally, metabolic rate can be calculated by times an oxygen consumption $\dot{V}_{o_2}$ (L/hr) at STPD to a caloric equivalent of oxygen (Cal/hr) divided by Body Surface Area (BSA) in m^2^, which can be calculated by [DuBois & DuBois formula](http://www-users.med.cornell.edu/~spon/picu/calc/bsacalc.htm) (DuBois D, DuBois EF)



<!-- $$ -->
<!-- Met \ Rate \ (Cal/m^2/hr) = \frac{ \dot{V}_{o_2} (L/hr)  \times CalEqi \ O_2 \ (Cal/hr)}{ BSA \ (m^2) }  -->
<!-- $$ -->

<!-- ![equation](https://latex.codecogs.com/svg.latex?Met%20%5C%20Rate%20%5C%20%28Cal/m%5E2/hr%29%20%3D%20%5Cfrac%7B%20%5Cdot%7BV%7D_%7Bo_2%7D%20%28L/hr%29%20%5Ctimes%20CalEqi%20%5C%20O_2%20%5C%20%28Cal/hr%29%7D%7B%20BSA%20%5C%20%28m%5E2%29%20%7D) -->

\

$$
Met \ Rate \ (Cal/m^2/hr) = \frac{ \dot{V}_{o_2} (L/hr)  \times CalEqi \ O_2 \ (Cal/hr)}{ BSA \ (m^2) }
$$
\




`get_metabolic_rate()` is a final wrapper function that calculate metabolic rate; moreover, it also reports metabolic rate along with oxygen consumption and other related parameters printed to the R console.

```{r}
get_metabolic_rate(x = 15, # displacement in x-direction = 15 mm
                   y = 80, # displacement in y-direction = 80 mm
                   paper_speed = 25, # Paper speed of the kymograph = 25 mm/min
                   baro = 760, # Barometric pressure at the recording site.
                   temp_c = 25, # Temperature in celsius at the recording site.
                   wt_kg = 70, # Subject's weight in kilogram
                   ht_cm = 180, # Subject's height in centimetre
                   cal_eqi_oxygen = 4.825 # Caloric equivalent of Oxygen at RQ = 0.82
)
```

            
            
            
            




## Simulation of Harvard Spirometer Tracing

### Simulate Data

In this section, contrary to the previous, we will simulate data to plot Harvard spirometer tracing from the respiratory parameters.

\

`sim_Harvard_tracing()` is the function to simulates volume-time tracing data produced by breathing of a hypothetical subject as recorded by the Harvard spirometer. 


The input parameters can be categorized into the followings:

- **Function to Simulate 1 Respiratory Cycle:** as supply by the argument `f`, the default, currently, is a cosine (`cos`) function which might give a rough representation of 1 respiratory cycle. (It is not a physiologic representation of respiratory waveform. I still need more research on this.)

- **Amplitude and Wavelength of the sinusoidal function:** this can be calculated by knowing the Respiratory Rate (`RR`), Tidal Volume (`TV`), and paper speed (`paper_speed`)

- **Slope and Y-intercept of an Oxygen Line:** the oxygen line, i.e., the linear line fitted from the lowest points of each respiratory waveform, can be simulated by knowing its slope ($\beta_1$), y-intercept ($\beta_0$), and random error term ($\epsilon$).

    - **Slope ($\beta_1$)** can be calculated from an oxygen consumption (argument `oxycons`) and paper speed (`paper_speed`).
    - **Y-intercept ($\beta_0$)** can be specified directly from the user.
    - And, **a random error ($\epsilon$)** can be generated from a Gaussian distribution with $\mu$ = 0, and `sd` as specified.

```{r tracing_df}
tracing_df <- sim_Harvard_tracing(
  f = "cos",         # Cosine function to represent 1 respiratory cycle
  t_start = 0,       # First `x` value will be started at time = 0 minute
  t_end = 1,         # Last `x` value will be ended at time = 1 minute
  paper_speed = 25,  # Paper speed in mm/minute
  y_int_O2_line = 5, # Y-intercept of an oxygen line
  oxycons = 10,      # Oxygen consumption is 10 L/hr (default unit)
  TV = 500,          # Tidal Volume is 500 ml
  RR = 20,           # Respiratory Rate is 20 /min
  seq_x_by = 0.01,   # This control resolution of simulated data
  epsilon_sd = 0.3,  # Add random variation sampled from Gaussian distribution with mean = 0, sd = 0.3
  seed = 1,          # Seed to generate the random variation
)

head(tracing_df)
```

### Plot

With a little help of `{ggplot2}` package, the tracing can then be visualized.

```{r}
library(ggplot2)
library(latex2exp)

theme_set(theme_bw())
```



```{r}
ggplot(tracing_df) + 
  geom_path(aes(x, y)) + 
  geom_smooth(aes(x, y_O2_line), method = "lm", formula = "y~x",
              linetype = "dashed") +
  expand_limits(y = 0) +
  labs(title = "Simulation of Harvard Spirometer Tracing",
        subtitle = TeX("(TV = 500 ml, RR = 20/min, $\\dot{V}_{O_2}$ = 10 L/hr)"),
        x = "x (mm)", y = "y (mm)")
```


## Math Behind the Simulation of Harvard Spirometer Tracing

To be more explicit, I will provide an underlying mathematical formula that produces data to simulate Harvard spirometer tracing in the following sections.

\

The goal here is to build a function that yields an output:

- $y$ = y-axis values (mm)

from inputs:

- $x$ = x-axis values (mm)

- $V_T$ = Tidal Volumes (ml)
- $f_R$ = Respiratory Rate (cycle/min)
- $\dot{V}_{O_2}$ = Oxygen consumption (L/hr)
- $v_{paper}$ = Paper Speed (mm/min)

\

$y$ can be decomposed into the sum of its 2 components: the respiratory waveform ($y_{wave}$) and the linear components ($y_{lin}$) which is the oxygen line.

$$
y = y_{wave} + y_{lin}
$$

Next, I will provide methods to build $y_{wave}$ and then $y_{lin}$.


### The Respiratory Waveform ($y_{wave}$)

For simplicity, let's assume that 1 respiratory wave can be generated by using trigonometric sinusoidal function. In this case I will use the cosine function `cos()`.

\

Thus, the *respiratory waveform* can be written as:

$$
y_{wave} = A \times cos(\frac{2\pi   x}{\lambda})
$$

Where:

- $A$ = amplitude of wave (mm)
- $\lambda$ = wave length (mm)
- $x$ = x-axis values (mm)

Given that 1 mm in the y-axis = 30 ml. and the amplitude ($A$) is half of tidal volume ($V_T$), 

Amplitude ($A$) can be expressed in terms of tidal volume ($V_T$) in ml.

$$
A = \frac{-TV}{2 \times 30}
$$

$2$ was included in the denominator because amplitude equals the half of tidal volume, and the minus sign was added to flip the cosine function horizontally.





