---
title: "Introduction to rslab"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to rslab}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



## BTPS Calculator

[Spirometry test](https://en.wikipedia.org/wiki/Spirometry) is a procedure that measures lung volumes and air flow parameters which calculate from inspiratory and expiratory gas volumes of a human subject.

The gas volumes/flows obtained by the spirometer are in **ATPS** (Ambient Temperature and Pressure Saturated) conditions which its volumes also depends on environmental conditions including room temperature.

The obtained volumes/flows must be converted to gas condition in **BTPS** (Body Temperature, Pressure, water vapor Saturated) which gives more accurate representation of actual volumes/flows within the lungs.

\

[BTPS correction factor](https://nddmed.com/_Resources/Persistent/8a716ac3fd123ce0becd7b56596582d4fc4c0c47/appnote-btps-correction-v01r.pdf) is a coefficient used to convert flow and volume measured at ambient conditions (ATPS) to the conditions within the lungs (BTPS).

`get_btps_factor()` is a function to obtain the BTPS correction factor.

```{r setup}
library(rslab)
```

```{r btps}
# BTPS correction factor at room temperature = 27 degree celsius
btps_factor27 <- get_btps_factor(temp = 27)
btps_factor27
```

If Tidal Volume (TV) was measured as 500 ml at room temperature of 27 degree celsius, the 
TV in the lung at BTPS (body temperature, pressure, water vapor saturated) would be:

```{r}
# TV (ml) at BTPS
500 * btps_factor27 
```

If you want to convert several lung volumes or flow (from ATPS to BTPS), use `lung_vol_atps_btps()` and the result will be shown as a tibble.

```{r lung_vol_conv}
lung_vol_atps_btps(
  temp = 27,
  FEV1 = 3.5, 
  FVC = 4.5,
  PEF = 450,
  TV = 0.5,
  IC = 2.5,
  EC = 2.5,
  VC = 4.5
)
```

## Metabolic Rate & Oxygen Consumption Calculator

In a laboratory experiment using [Harvard Spirometer](https://www.somatco.com/Recording-Spirometer-50-1833-50-1817.pdf), this package provides functions to calculate metabolic rate and oxygen consumption by input a displacement in x-and y-direction of Harvard spirometer tracing, height and weight of the subject, and environmental conditions which are temperature and barometric pressure.

### Oxygen Consumption

In Harvard spirometer tracing, horizontal direction (x-direction) represents time (default paper speed = 25 mm/min) whereas vertical direction (y-direction) represents usage of oxygen (1 mm = 30 ml of oxygen). With a little bit of calculation, we can derived an oxygen consumption in unit of L/hr in the **ATPS** condition (**A**mbient **T**emperature and barometric **P**ressure **S**aturated with water vapor condition).

You can use `get_oxycons()` to calculate oxygen consumption and it will also provide other info printed to R console as well. 

```{r oxycons}
oxycons <- get_oxycons(x = 15, # displacement in x-direction = 15 mm
            y = 80, # displacement in y-direction = 80 mm
            paper_speed = 25 # Paper speed of the kymograph = 25 mm/min
            )
```


### Metabolic Rate

To calculate metabolic rate, we must first convert oxygen consumption $\dot{V}_{o_2}$ at **ATPS** to **STPD** (**S**tandard **T**emperature of 0°C and a barometric **P**ressure of 760 mmHg, and in a **D**ry state).

STPD correction factor can be used to convert gas in ATPS to BTPS condition.
It has a linear relationship with barometric pressure and temperature. 

Therefore, `get_STPD_factor()` can be used to predict STPD factor using `baro` and `temp_c` as predictors. (It use multiple linear regression model behind the scene.)


```{r stpd_760_25}
stpd_760_25 <- get_STPD_factor(
  baro = 760, # Barometric pressure at the recording site.
  temp_c = 25 # Temperature in celsius at the recording site.
  ) 
```

Correction can be made by multiply oxygen consumption $\dot{V}_{o_2}$ at ATPS to STPD correction factor.

```{r}
oxycons * stpd_760_25 # Unit in L/hr
```

Finally, metabolic rate can be calculated by times an oxygen consumption $\dot{V}_{o_2}$ (L/hr) at STPD to a caloric equivalent of oxygen (Cal/hr) divided by Body Surface Area (BSA) in m^2^, which can be calculated by [DuBois & DuBois formula](http://www-users.med.cornell.edu/~spon/picu/calc/bsacalc.htm) (DuBois D, DuBois EF)



<!-- $$ -->
<!-- Met \ Rate \ (Cal/m^2/hr) = \frac{ \dot{V}_{o_2} (L/hr)  \times CalEqi \ O_2 \ (Cal/hr)}{ BSA \ (m^2) }  -->
<!-- $$ -->

<!-- ![equation](https://latex.codecogs.com/svg.latex?Met%20%5C%20Rate%20%5C%20%28Cal/m%5E2/hr%29%20%3D%20%5Cfrac%7B%20%5Cdot%7BV%7D_%7Bo_2%7D%20%28L/hr%29%20%5Ctimes%20CalEqi%20%5C%20O_2%20%5C%20%28Cal/hr%29%7D%7B%20BSA%20%5C%20%28m%5E2%29%20%7D) -->

\

$$
Met \ Rate \ (Cal/m^2/hr) = \frac{ \dot{V}_{o_2} (L/hr)  \times CalEqi \ O_2 \ (Cal/hr)}{ BSA \ (m^2) }
$$
\


`get_metabolic_rate()` is a final wrapper function that calculate metabolic rate; moreover, it also reports metabolic rate along with oxygen consumption and other related parameters printed to the R console.

```{r}
get_metabolic_rate(x = 15, # displacement in x-direction = 15 mm
                   y = 80, # displacement in y-direction = 80 mm
                   paper_speed = 25, # Paper speed of the kymograph = 25 mm/min
                   baro = 760, # Barometric pressure at the recording site.
                   temp_c = 25, # Temperature in celsius at the recording site.
                   wt_kg = 70, # Subject's weight in kilogram
                   ht_cm = 180, # Subject's height in centimetre
                   cal_eqi_oxygen = 4.825 # Caloric equivalent of Oxygen at RQ = 0.82
)
```

            
            
            
            




