---
title: "Introduction to rslab"
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig.align: 'center'
    fig.asp: 0.618
    dpi: 320
    toc: true
    warning: FALSE
    message: FALSE
vignette: >
  %\VignetteIndexEntry{Introduction to rslab}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  dpi = 300,
  warning = FALSE,
  message = FALSE,
  out.width = "100%",
  comment = "#>"
)
```







## BTPS Calculator

[Spirometry test](https://en.wikipedia.org/wiki/Spirometry) is a procedure that measures lung volumes and air flow parameters which calculate from inspiratory and expiratory gas volumes of a human subject.

The gas volumes/flows obtained by the spirometer are in **ATPS** (Ambient Temperature and Pressure Saturated) conditions which its volumes also depends on environmental conditions including room temperature.

The obtained volumes/flows must be converted to gas condition in **BTPS** (Body Temperature, Pressure, water vapor Saturated) which gives more accurate representation of actual volumes/flows within the lungs.

\

[BTPS correction factor](https://nddmed.com/_Resources/Persistent/8a716ac3fd123ce0becd7b56596582d4fc4c0c47/appnote-btps-correction-v01r.pdf) is a coefficient used to convert flow and volume measured at ambient conditions (ATPS) to the conditions within the lungs (BTPS).

`get_btps_factor()` is a function to obtain the BTPS correction factor.

```{r setup}
library(rslab)
```

```{r btps}
# BTPS correction factor at room temperature = 27 degree celsius
btps_factor27 <- get_btps_factor(temp = 27)
btps_factor27
```

If Tidal Volume (TV) was measured as 500 ml at room temperature of 27 degree celsius, the 
TV in the lung at BTPS (body temperature, pressure, water vapor saturated) would be:

```{r}
# TV (ml) at BTPS
500 * btps_factor27 
```

If you want to convert several lung volumes or flow (from ATPS to BTPS), use `lung_vol_atps_btps()` and the result will be shown as a tibble.

```{r lung_vol_conv}
lung_vol_atps_btps(
  temp = 27,
  FEV1 = 3.5, 
  FVC = 4.5,
  PEF = 450,
  TV = 0.5,
  IC = 2.5,
  EC = 2.5,
  VC = 4.5
)
```

## Metabolic Rate & Oxygen Consumption Calculator

In a laboratory experiment using [Harvard Spirometer](https://www.somatco.com/Recording-Spirometer-50-1833-50-1817.pdf), this package provides functions to calculate metabolic rate and oxygen consumption by input a displacement in x-and y-direction of Harvard spirometer tracing, height and weight of the subject, and environmental conditions which are temperature and barometric pressure.

### Oxygen Consumption

In Harvard spirometer tracing, horizontal direction (x-direction) represents time (default paper speed = 25 mm/min) whereas vertical direction (y-direction) represents usage of oxygen (1 mm = 30 ml of oxygen). With a little bit of calculation, we can derived an oxygen consumption in unit of L/hr in the **ATPS** condition (**A**mbient **T**emperature and barometric **P**ressure **S**aturated with water vapor condition).

You can use `get_oxycons()` to calculate oxygen consumption and it will also provide other info printed to R console as well. 

```{r oxycons}
oxycons <- get_oxycons(x = 15, # displacement in x-direction = 15 mm
            y = 80, # displacement in y-direction = 80 mm
            paper_speed = 25 # Paper speed of the kymograph = 25 mm/min
            )
```


### Metabolic Rate

To calculate metabolic rate, we must first convert oxygen consumption $\dot{V}_{o_2}$ at **ATPS** to **STPD** (**S**tandard **T**emperature of 0°C and a barometric **P**ressure of 760 mmHg, and in a **D**ry state).

STPD correction factor can be used to convert gas in ATPS to BTPS condition.
It has a linear relationship with barometric pressure and temperature. 

Therefore, `get_STPD_factor()` can be used to predict STPD factor using `baro` and `temp_c` as predictors. (It use multiple linear regression model behind the scene.)


```{r stpd_760_25}
stpd_760_25 <- get_STPD_factor(
  baro = 760, # Barometric pressure at the recording site.
  temp_c = 25 # Temperature in celsius at the recording site.
  ) 
```

Correction can be made by multiply oxygen consumption $\dot{V}_{o_2}$ at ATPS to STPD correction factor.

```{r}
oxycons * stpd_760_25 # Unit in L/hr
```

Finally, metabolic rate can be calculated by times an oxygen consumption $\dot{V}_{o_2}$ (L/hr) at STPD to a caloric equivalent of oxygen (Cal/hr) divided by Body Surface Area (BSA) in m^2^, which can be calculated by [DuBois & DuBois formula](http://www-users.med.cornell.edu/~spon/picu/calc/bsacalc.htm) (DuBois D, DuBois EF)



<!-- $$ -->
<!-- Met \ Rate \ (Cal/m^2/hr) = \frac{ \dot{V}_{o_2} (L/hr)  \times CalEqi \ O_2 \ (Cal/hr)}{ BSA \ (m^2) }  -->
<!-- $$ -->

<!-- ![equation](https://latex.codecogs.com/svg.latex?Met%20%5C%20Rate%20%5C%20%28Cal/m%5E2/hr%29%20%3D%20%5Cfrac%7B%20%5Cdot%7BV%7D_%7Bo_2%7D%20%28L/hr%29%20%5Ctimes%20CalEqi%20%5C%20O_2%20%5C%20%28Cal/hr%29%7D%7B%20BSA%20%5C%20%28m%5E2%29%20%7D) -->

\

$$
Met \ Rate \ (Cal/m^2/hr) = \frac{ \dot{V}_{o_2} (L/hr)  \times CalEqi \ O_2 \ (Cal/hr)}{ BSA \ (m^2) }
$$
\




`get_metabolic_rate()` is a final wrapper function that calculate metabolic rate; moreover, it also reports metabolic rate along with oxygen consumption and other related parameters printed to the R console.

```{r}
get_metabolic_rate(x = 15, # displacement in x-direction = 15 mm
                   y = 80, # displacement in y-direction = 80 mm
                   paper_speed = 25, # Paper speed of the kymograph = 25 mm/min
                   baro = 760, # Barometric pressure at the recording site.
                   temp_c = 25, # Temperature in celsius at the recording site.
                   wt_kg = 70, # Subject's weight in kilogram
                   ht_cm = 180, # Subject's height in centimetre
                   cal_eqi_oxygen = 4.825 # Caloric equivalent of Oxygen at RQ = 0.82
)
```

            
            
            
            




## Simulation of Harvard Spirometer Tracing

### Simulate Data

In this section, contrary to the previous, we will simulate data to plot Harvard spirometer tracing from the respiratory parameters.

\

`sim_Harvard_tracing()` is the function to simulates volume-time tracing data produced by breathing of a hypothetical subject as recorded by the Harvard spirometer. 


The input parameters can be categorized into the followings:

- **Function to Simulate 1 Respiratory Cycle:** as supply by the argument `f`, the default, currently, is a cosine (`cos`) function which might give a rough representation of 1 respiratory cycle. (It is not a physiologic representation of respiratory waveform. I still need more research on this.)

- **Amplitude and Wavelength of the sinusoidal function:** this can be calculated by knowing the Respiratory Rate (`RR`), Tidal Volume (`TV`), and paper speed (`paper_speed`)

- **Slope and Y-intercept of an Oxygen Line:** the oxygen line, i.e., the linear line fitted from the lowest points of each respiratory waveform, can be simulated by knowing its slope ($\beta_1$), y-intercept ($\beta_0$), and random error term ($\epsilon$).

    - **Slope ($\beta_1$)** can be calculated from an oxygen consumption (argument `oxycons`) and paper speed (`paper_speed`).
    - **Y-intercept ($\beta_0$)** can be specified directly from the user.
    - And, **a random error ($\epsilon$)** can be generated from a Gaussian distribution with $\mu$ = 0, and `sd` as specified.

```{r tracing_df}
tracing_df <- sim_Harvard_tracing(
  f = "cos",         # Cosine function to represent 1 respiratory cycle
  t_start = 0,       # First `x` value will be started at time = 0 minute
  t_end = 1,         # Last `x` value will be ended at time = 1 minute
  paper_speed = 25,  # Paper speed in mm/minute
  y_int_O2_line = 5, # Y-intercept of an oxygen line
  oxycons = 10,      # Oxygen consumption is 10 L/hr (default unit)
  TV = 500,          # Tidal Volume is 500 ml
  RR = 20,           # Respiratory Rate is 20 /min
  seq_x_by = 0.01,   # This control resolution of simulated data
  epsilon_sd = 0.3,  # Add random variation sampled from Gaussian distribution with mean = 0, sd = 0.3
  seed = 1,          # Seed to generate the random variation
)

head(tracing_df)
```

### Plot

With a little help of `{ggplot2}` package, the tracing can then be visualized.

```{r}
library(ggplot2)
library(latex2exp)

theme_set(theme_bw())
```



```{r}
ggplot(tracing_df) + 
  geom_path(aes(x, y)) + 
  geom_smooth(aes(x, y_O2_line), method = "lm", formula = "y~x",
              linetype = "dashed") +
  expand_limits(y = 0) +
  labs(title = "Simulation of Harvard Spirometer Tracing",
        subtitle = TeX("(TV = 500 ml, RR = 20/min, $\\dot{V}_{O_2}$ = 10 L/hr)"),
        x = "x (mm)", y = "y (mm)")
```

## The Modelling of Harvard Spirometer Tracing

The underlying mathematical models that produces data to simulate Harvard spirometer tracing will be explained in the following sections.

\

Since the aim is to build a 2D plot with x-and y-position in milimeter, it requires a function to generate y-axis data from x-axis data and other physiologic parameters.

Therefore, the goal here is to build a function $f_{tracing}(x)$ that yields an output:

- $y$ = y-axis values (mm)

from inputs:

- $x$ = x-axis values (mm)

and physiologic parameters:

- $V_T$ = Tidal Volumes (ml)
- $f_R$ = Respiratory Rate (cycle/min)
- $\dot{V}_{O_2}$ = Oxygen consumption (L/hr)
- $v_{paper}$ = Paper Speed (mm/min)

\


$f_{tracing}(x)$ can be decomposed into the sum of its 2 components: the respiratory waves function $f_{wave}(x)$ and the oxygen line function $f_{O_2line}$ that results from oxygen being consumed by the experimental subject (if the oxygen consumption ($\dot{V}_{O_2}$) is constant).


<!-- $f(x)$ can be decomposed into the sum of its 2 components: the respiratory waves function $f_{wave}(x)$ and the linear function $f_{lin}(x)$ which produces the oxygen line (assuming that oxygen consumption ($\dot{V}_{O_2}$) is constant). -->

<!-- $y$ can be decomposed into the sum of its 2 components: the respiratory waveform ($y_{wave}$) and the linear components ($y_{lin}$) which is the oxygen line. -->

\

$$
f_{tracing}(x) = f_{wave}(x) + f_{O_2line}(x)
$$

\

Next, I will provide methods to build $f_{wave}(x)$ and then $f_{O_2line}(x)$.


### The Respiratory Waves Function

For simplicity, respiratory waves function $f_{wave}(x)$ will be modeled from trigonometric sinusoidal function. In this case, I will use the cosine function `cos()`.

The *general form* of cosine function that models position $y$ as a function of the position $x$ can be written as:

\

$$
y(x) = A \times cos(k x) \\
k = \frac{2\pi}{\lambda}
$$


Where:

- $A$ = the amplitude of wave (mm)
- $k$ = the wave number (1/mm)
- $\lambda$ = wave length (mm)

\

To model respiratory waves,


the **amplitude of wave** ($A$) in mm can be expressed in terms of *tidal volume* ($V_T$) in ml, given that 1 mm in the y-axis = 30 ml.


$$
A = \frac{V_T}{2 \times 30}
$$

<!-- (The minus sign was added to flip the cosine function horizontally which results in an upward bell curve.) -->

\

And, the **wave length** ($\lambda$) in mm can be expressed in terms of *respiratory rate* ($f_R$) in cycle/min and paper speed ($v_{paper}$) in mm/min.

$$
\lambda = \frac{v_{paper}}{f_R} 
$$


Therefore, the **respiratory waves function** $f_{wave}(x)$ can be written as:

\


$$
f_{wave}(x) = \frac{V_T}{60} \times cos(\frac{2\pi f_R}{v_{paper}} \times x)
$$
\

Where:

- $V_T$ = Tidal Volumes (ml)
- $f_R$ = Respiratory Rate (cycle/min)
- $v_{paper}$ = Paper Speed (mm/min)



### The Oxygen Line Function

Assuming that the rate of oxygen used by the subject is constant, the function that represents oxygen consumption $f_{O_2line}$ could be modeled by the linear equation.

Since The simple linear equation is expressed in the following form:

$$
y(x) = \beta_0 + \beta_1 x
$$

Where:

- $\beta_0$ = y-intercept 
- $\beta_1$ = slope

Modelling slope ($\beta_1$) in terms of the oxygen consumption ($\dot{V}_{O_2}$) in L/hr and paper speed ($v_{paper}$) in mm/min can be written as follows.

$$
\beta_1 = \frac{ \dot{V}_{O_2} }{ v_{paper} \times 30 } \times \frac{1000}{60} = \frac{5  \dot{V}_{O_2} }{ 9 v_{paper} }
$$


Therefore, the linear function $f_{O_2line}$ which produces the oxygen line can be written as:

\

$$
f_{O_2line}(x) = \beta_0 + \frac{5  \dot{V}_{O_2} }{ 9 v_{paper} } \times x
$$
\

Where:

- $\beta_0$ = y-intercept (mm)
- $\dot{V}_{O_2}$ = Oxygen consumption (L/hr)
- $v_{paper}$ = Paper Speed (mm/min)



### The Harvard Spirometer Tracing Function

We have constructed the respiratory waves function $f_{wave}(x)$:

\

$$
f_{wave}(x) = \frac{V_T}{60} \times cos(\frac{2\pi f_R}{v_{paper}} \times x)
$$
\

and, the oxygen line function $f_{O_2line}$.

\

$$
f_{O_2line}(x) = \beta_0 + \frac{5  \dot{V}_{O_2} }{ 9 v_{paper} } \times x
$$

Since, the function that produces Harvard spirometer tracing $f_{tracing}(x)$ can be written as the sum of $f_{wave}(x)$ and $f_{O_2line}$:

$$
f_{tracing}(x) = f_{wave}(x) + f_{O_2line}(x) 
$$


substituting $f_{wave}(x)$ and $f_{O_2line}(x)$ with above equations yields the final equation that produce Harvard spirometer tracing $f_{tracing}(x)$.

\

$$
y(x) = f_{tracing}(x) = \frac{V_T}{60} . cos(\frac{2\pi x f_R }{v_{paper}}) + \frac{5 x \dot{V}_{O_2} }{ 9 v_{paper} } + \beta_0 
$$

\

Where:

- Axis values:

    - $y$ = y-axis values (mm)
    - $x$ = x-axis values (mm)

- Physiologic parameters:

    - $V_T$ = Tidal Volumes (ml)
    - $f_R$ = Respiratory Rate (cycle/min)
    - $\dot{V}_{O_2}$ = Oxygen consumption (L/hr)
    - $v_{paper}$ = Paper Speed (mm/min)
    - $\beta_0$ = y-intercept 


